### 2020-03-03 main_text/main_text.tex
Implemented some of GC's edits

	It is quite conceivable that this auto-activation process is common to many
	substances that a cell would want to accumulate. Such a switch enables the
	cell to sense and respond to its environment: if the substrate enters the
	cell, it activates the production of membrane transporters. The cell then
	starts accumulating the substrate, thereby ``testing'' the substrate's
	presence in the extracellular environment. If there is enough, the
	expression stabilizes at an ``on'' state and the cell has, in a short-term
	sense, adapted. When there is not enough substrate, the operon, which
	often encodes for a whole set of enzymes for this one metabolite, switches
	``off'' again. Such a mechanism could be involved in various cases of
	short-term adaptation.

	A key element of this mechanism is the presence of a transcription factor
	which binds to the transporter substrate and which is often expressed at a low level (often at copy numbers
	of order $\sim10$, \cite{Schmidt2015}).
	This is resource efficient for the cell, as this low copy number
	transcription factor acts as an ``always on'' sensor to detect
	the substrate, allowing high copy numbers of the membrane transporter
	and its attendant operon to be expressed only when
	their substrate is actually present. The transcription factors
	LacI, AraC, and XylR all appear to fill this
	role~\cite{Novick1957,Santillan2008,Ozbudak2004,Narang2008,Choi2008,Fritz2014,Jenkins2017,Siegele1997,Song1997,Schmidt2015}.

@@

	\section*{Experimental motivation} \label{sec:experiment}
	Our work was motivated by the experimental observation of bimodality
	in the \emph{xap} circuit, which is shown in Fig~\ref{fig1:data}. We
	focus on the essential findings here and refer the reader
	to~\nameref{S1_Text} for more experimental details. Briefly, we
	placed a fluorescent reporter under the control of the wild type
	(wt) \emph{xapAB} promoter. This construct was placed in three
	different backgrounds: \textDelta\emph{xapABR},
	\textDelta\emph{xapAB}, and wt. The left panel of
	Fig~\subref{fig1:data}{A} shows that for increasing xanthosine
	concentration, expression in the wt-background increases in a
	switch-like way: it is nearly zero when there is little xanthosine
	(``uninduced state''), but increases drastically when there is more
	(``induced state''). In between, the aforementioned bimodal
	distribution is obtained, where some bacteria are in the uninduced
	expression state and others are induced.

	\begin{figure}%[h!]
		\centering
		\includegraphics[width=1\textwidth]{media/Fig1_experiment.eps}
		\caption{{\bf Experimental data on the \emph{xap} circuit.}
		(A) The expression of the \emph{xapAB} promoter was measured for
		different extracellular concentrations of xanthosine.
		The left panel shows the wild type circuit while
	in the middle the genes \emph{xapABR} and on the right the genes
		\emph{xapAB} are removed. The wild type circuit behaves like a
		switch. Note that the fluorescence scale of the middle panel
		is not comparable with the other two,
		and also that the chosen xanthosine concentrations are different.
		(B) shows the fold-change in gene expression upon addition of
		xanthosine for the wt promoter (left panel)
		and for the promoter with only the XapR binding site adjacent
		to the polymerase binding site (right panel).
		Note that the fold-change used here differs from fold-change in,
		e.g.,~\cite{Garcia2011,RazoMejia2018,Chure2019},
		in that no subtraction of autofluorescence was carried out,
		which is adequate for the qualitative comparison
		of these two promoters.
		}
		\label{fig1:data}
	\end{figure}

@@

	Can we gain further insight into how XapR regulates
	the \emph{xapAB} promoter? As mentioned in
	the introduction, there are two binding sites for XapR on the
	\emph{xapAB} promoter: one directly at the polymerase binding site
	and one further upstream.
	Working in the \textDelta\emph{xapAB} background,
	we measured the expression level of our reporter when driven by two
	constructs: the wild-type \emph{xapAB} promoter
	(Fig~\subref{fig1:data}{B}, left panel), and the \emph{xapAB}
	promoter with the upstream binding site removed
	(Fig~\subref{fig1:data}{B}, right panel).
	Clearly, removing one XapR binding site dramatically reduced the
	responsiveness of the promoter to xanthosine (though not apparent
	from Fig~\subref{fig1:data}{B}, this response remains detectable).

@@

	\section*{Model}
	\subsection*{Step by step modeling of the system}
	In this section, we present our model of the \emph{xapAB} genetic circuit.
	Fig~\ref{fig2:model} shows an overview of this model.
	The qualitative picture of the circuit switching its state is as follows:
	\begin{itemize}
		\item In the initial absense of XapB, small amounts of xanthosine permeate
				into and out of the cell (discussed in more detail below).
		\item XapR is induced by xanthosine.
		\item The induced XapR binds to the \emph{xapAB} promoter, leading to transcription.
		\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
		\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
		\item Production of more XapA and XapB is balanced by decay or dilution through cell divisions.
	\end{itemize}
	Because xanthosine induces the transcription factor XapR, we have positive
	and negative feedback loops due to XapB and XapA, respectively. The remainder
	of this subsection discusses each of the above steps in detail, leading us
	to a set of two coupled ODEs. More in-depth explanations can be found
	in~\nameref{S1_Text}.

@@

	With $\n{[m]}$ being the mRNA concentration, $r_{\n{m}}$ the transcription
	rate, $\gamma_{\n{m}}$ the mRNA decay rate, and $p_{\n{active}}$ the
	probability of the promoter being in the active state, we obtain
	\begin{eqnarray}
	\label{eq:mbasic}
	\dd{\n{[m]}}{t} &=& r_{\n{m}} p_{\n{active}} - \gamma_{\n{m}} \n{[m]} \\
	p_{\n{active}} &=& \frac{w_8}{\sum_{i=1}^{8} w_i} = 
	\frac{\n{[P]}}{K_{\n{P}}+\n{[P]}} 
	\frac{
		\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 
		e^{- \beta \Delta E_{\n{coop}}}
	}{
		1 + 
		2 \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} +
		\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 e^{- \beta \Delta E_{\n{coop}}}
	}
	\end{eqnarray}
	Here, $w_i$ stands for the thermodynamic weight of the ith state in the
	order in which they are listed in Fig~\ref{fig3:states}. As written above,
	the partition function factorizes into a polymerase and a XapR term because
	of our assumption of independent binding, which is further discussed
	in~\nameref{S1_Text}. Note that because $r_{\n{m}}$ implicitly contains the
	gene copy number per cell, it has units of $\n{M^{-1} \ s^{-1}}$ and not
	just $\n{s^{-1}}$. This rate equation gives the mean mRNA concentration
	$\langle\n{[mRNA]}\rangle = \frac{r_{\n{m}}}{\gamma_{\n{m}}}
	p_{\n{active}}$, which we will need in the next paragraph. The mean can also be found from the full chemical master equation, which is shown in~\nameref{S1_Text}.

@@

	\subsection*{Deterministic phase portraits}
	A standard way to analyze dynamical systems deterministically is
	to plot phase portraits. In the following, we present such plots where the
	state variables are the mRNA, the protein, and the xanthosine concentration. Note that for the low copy numbers that can occur in our system, a deterministic analysis is not necessarily meaningful. Nevertheless, in our case we find that stochastic simulations agree well with the deterministic results. Thus, deterministic phase portraits are a surprisingly good starting point.

@@

	As a remark we point out that we have not observed any oscillations in the
	system. Intuitively, they might be expected when the XapA rate is
	significantly larger than the XapB rate, but it turns out that oscillations
	cannot be obtained. Why they do not occur can be understood when looking at the regions that
	are bounded by all three nullclines: on these boundaries, the streamlines
	point into the bounded regions, so deterministically, they serve as
	trapping regions from which the system cannot escape. Once inside, the only
	possible trajectory is non-oscillatory flow towards the stable fixed point.

@@

	\paragraph*{The roles of XapA and XapB.}
	While it is clear that the bistability in the model system is due to the
	feedback loop from XapA and XapB, it is not intuitively clear if both XapA and XapB
	are necessary. The model implies that the bistability is due to XapB only. XapA
	is neither sufficient nor necessary and, within the estimated parameter
	regime, does not even have a significant influence on the system. This can
	be seen from the plots in figure~\ref{fig6:xapAB}.
	Degradation of xanthosine by XapA lowers the xanthosine and protein concentration at the upper
	fixed point by a small amount and could, in principle, thereby make the
	high-expression solution vanish. For our choice of all other parameters,
	bistability only vanishes for $k_{\n{\alpha}} > 10^4$ which is far from what has been
	measured. However, a higher effective rate could, in principle, be achieved by
	different translation rates of XapA and XapB (see simplifications of the
	model in~\nameref{S1_Text}). Hence, we cannot exclude the possibility that
	XapA becomes so strong that it makes bistability impossible, but this is an
	extreme case. XapB, on the other hand, is essential; without it the system
	only has the one fixed point at low expression.\footnote{
		Seeger et.\ al.\cite{Seeger1995} observed that $\Delta xapB$ mutants could
		survive, but grew extremely slowly, with xanthosine as the only carbon
		source, which makes sense in light of Fig~\ref{fig6:xapAB}.
		With \textit{xapB} removed, the switch never activates and the cells are
		forced to survive with an extremely meager quantity of XapA to metabolize
		the abundant xanthosine.}

@@

	\paragraph*{The role of cooperativity.} 
	The model has two cooperatively interacting binding sites for XapR on the
	\emph{xapAB} promoter and two cooperative binding sites on XapR for
	xanthosine. It is interesting to consider whether the cooperativity is
	a necessary feature for bistability. This question is motivated by the importance of
	cooperativity in ``typical'' genetic switches\cite{Gardner2000,Cherry2000}.

@@

	\paragraph*{Bimodality and the extracellular xanthosine concentration.}
	The stochastic approach results in the same bimodal distributions that were already
	seen in the deterministic investigation and experimental studies.
	Fig~\ref{fig8:stochC} shows the distribution of protein expression found in
	the simulations for different values of the extracellular xanthosine
	concentration. The parameters that were used are the same as in the previous
	section (listed in Table~\ref{table1:nondim}). However, we also find good agreement of the simulations and the deterministic results for other sets of parameters. To obtain the distributions,
	we ran the simulation 5000 times for a simulated time of $10^6 \unit{s}$
	each and started at a mRNA, protein and intracellular xanthosine count of 0.

@@

	The results agree very well with the deterministic fixed points and
	experiments: the mean numbers of mRNA, protein, and xanthosine in the
	stochastic results are as predicted from the phase portraits. It does,
	however, become clear that the phase portraits do not tell whether the cells
	will actually populate both the high and the low expression state,
	because they do not show the effective barrier height between the two states.
	In Fig~\subref{fig8:stochC}{(A)}, a deterministically bistable scenario is shown where the cells never switched to the high
	expression state during the run time of our simulations. This result
	implies that bistability, characterized by two separate
	\emph{stable} steady-states, largely persists even in the presence
	of stochastic fluctuations, suggesting that the deterministic
	picture is surprisingly and unreasonably effective. In other words,
	the fact that bimodality only occurs with some fine tuning of
	parameters means that the circuit is a strong switch and the
	deterministic picture is sufficient except in a small region of
	parameter space. Whether or not the system is actually in that
	region of parameter space remains a question for future experiments.

@@

	\section*{Acknowledgments}
	We thank Patrick Lenggenhager for his help with \emph{Mathematica} issues, Jane Kondev and Jin Wang for interesting discussions, and Nigel Orme for assistance with illustrations. \textcolor{red}{Portions of the experiments reported here were carried out at the
	Physiology Course at the Marine Biological Laboratory in Woods
	Hole, MA, operated by the University of Chicago.}
	This work was supported by 1R35 GM118043-01 Maximizing Investigators'
	Research Award (MIRA) (to R.P.), and the Werner Siemens Foundation
	through the Swiss Study Foundation (K.S.L.). This material is based upon
	work supported by the National Science Foundation Graduate Research
	Fellowship under Grant No.\ DGE-1745301 (M.J.M.).

>>>>

	It is quite conceivable that this auto-activation process is common to many
	substances that a cell would want to consume. Such a switch enables the
	cell to sense and respond to its environment: if the substrate enters the
	cell, it activates the production of membrane transporters. The cell then
	starts accumulating the substrate, thereby ``testing'' the substrate's
	presence in the extracellular environment. If there is enough, the
	expression stabilizes at an ``on'' state and the cell has, in a short-term
	sense, adapted. When there is not enough substrate, the operon, which
	often encodes for a whole set of enzymes for this one metabolite, switches
	``off'' again. Such a mechanism could be involved in various cases of
	short-term adaptation.

	A key element of this mechanism is the presence of a transcription factor
	which binds to the transporter substrate and which is often expressed at a low level (often at copy numbers
	of order $\sim10$, \cite{Schmidt2015, Li2014}).
	This is resource efficient for the cell, as this low copy number
	transcription factor acts as an ``always on'' sensor to detect
	the substrate, allowing high copy numbers of the membrane transporter
	and its attendant operon to be expressed only when
	their substrate is actually present. The transcription factors
	LacI, AraC, and XylR all appear to fill this
	role~\cite{Novick1957,Santillan2008,Ozbudak2004,Narang2008,Choi2008,Fritz2014,Jenkins2017,Siegele1997,Song1997,Schmidt2015}.

@@

	\section*{Experimental motivation} \label{sec:experiment}
	Our work was motivated by the experimental observation of bimodality
	in the \emph{xap} circuit, which is shown in Fig~\ref{fig1:data}. We
	focus on the essential findings here and refer the reader
	to~\nameref{S1_Text} for more experimental details. Briefly, we
	placed a fluorescent reporter under the control of the wild type
	(wt) \emph{xapAB} promoter. This construct was placed in three
	different backgrounds: \textDelta\emph{xapABR},
	\textDelta\emph{xapAB}, and wt,
	and expression as a function of extracellular xanthosine concentration
	was measured using flow cytometetry, as has been described
	previously~\cite{RazoMejia2018}. The left panel of
	Fig~\subref{fig1:data}{A} shows that for increasing xanthosine
	concentration, expression in the wt-background increases in a
	switch-like way: it is nearly zero when there is little xanthosine
	(``uninduced state''), but increases drastically when there is more
	(``induced state''). In between, the aforementioned bimodal
	distribution is obtained, where some bacteria are in the uninduced
	expression state and others are induced.

	\begin{figure}%[h!]
		\centering
		\includegraphics[width=1\textwidth]{media/Fig1_experiment.eps}
		\caption{{\bf Experimental data on the \emph{xap} circuit.}
		(A) The expression of the \emph{xapAB} promoter was measured for
		different extracellular concentrations of xanthosine (vertical axis).
		The left panel shows the wild type circuit while
		the middle and right panels show the effect of deleting
		the genes \emph{xapABR} and \emph{xapAB}, respectively.
		The wild type circuit behaves like a
		switch. Note that the fluorescence scale of the middle panel
		is not comparable with the other two,
		and also that the chosen xanthosine concentrations are different.
		(B) shows the fold-change in gene expression upon addition of
		xanthosine for the wt promoter (left panel)
		and for the promoter with only the XapR binding site adjacent
		to the polymerase binding site (right panel).
		Note that the fold-change used here differs from fold-change in,
		e.g.,~\cite{Garcia2011,RazoMejia2018,Chure2019},
		in that no subtraction of autofluorescence was performed,
		which is adequate for the qualitative comparison
		of these two promoters.
		}
		\label{fig1:data}
	\end{figure}

@@

	As mentioned in
	the introduction, there are two binding sites for XapR on the
	\emph{xapAB} promoter: one directly at the polymerase binding site
	and one further upstream.
	Working in the \textDelta\emph{xapAB} background,
	we measured the expression level of our reporter when driven by two
	constructs: the wild-type \emph{xapAB} promoter
	(Fig~\subref{fig1:data}{B}, left panel), and the \emph{xapAB}
	promoter with the upstream binding site removed
	(Fig~\subref{fig1:data}{B}, right panel).
	Clearly, removing one XapR binding site dramatically reduced the
	responsiveness of the promoter to xanthosine (though not apparent
	from Fig~\subref{fig1:data}{B}, this response remains detectable).

@@

	\section*{Model}
	\subsection*{Step by step modeling of the system}
	In this section, we present our model of the \emph{xapAB} genetic circuit.
	Fig~\ref{fig2:model} shows an overview of this model.
	The qualitative picture of the circuit switching its state is as follows:
	\begin{itemize}
		\item In the initial absense of XapB, small amounts of xanthosine permeate
				into and out of the cell (discussed in more detail below).
		\item The presence of xanthosine shifts XapR's equilibrium from inactive to active.
		\item Active XapR binds to the \emph{xapAB} promoter, leading to transcription.
		\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
		\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
		\item Production of more XapA and XapB is balanced by degradation or dilution through cell divisions.
	\end{itemize}
	Because xanthosine activates the transcription factor XapR, we have positive
	and negative feedback loops due to XapB and XapA, respectively. The remainder
	of this subsection discusses each of the above steps in detail, leading us
	to a set of two coupled ODEs. More in-depth explanations can be found
	in~\nameref{S1_Text}.

@@

	With $\n{[m]}$ being the mRNA concentration, $r_{\n{m}}$ the transcription
	rate, $\gamma_{\n{m}}$ the mRNA decay rate, and $p_{\n{active}}$ the
	probability of the promoter being in the active state, we obtain
	\begin{eqnarray}
	\label{eq:mbasic}
	\dd{\n{[m]}}{t} &=& r_{\n{m}} p_{\n{active}} - \gamma_{\n{m}} \n{[m]} \\
	p_{\n{active}} &=& \frac{w_8}{\sum_{i=1}^{8} w_i} = 
	\frac{\n{[P]}}{K_{\n{P}}+\n{[P]}} 
	\frac{
		\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 
		e^{- \beta \Delta E_{\n{coop}}}
	}{
		1 + 
		2 \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} +
		\left( \frac{\mathrm{[XapR]_A}}{K_{\mathrm{XapR}}} \right)^2 e^{- \beta \Delta E_{\n{coop}}}
	}
	\end{eqnarray}
	Here, $w_i$ stands for the thermodynamic weight of the ith state in the
	order in which they are listed in Fig~\ref{fig3:states}. As written above,
	the partition function factorizes into a polymerase and a XapR term because
	of our assumption of independent binding, which is further discussed
	in~\nameref{S1_Text}. Note that because $r_{\n{m}}$ implicitly contains the
	gene copy number per cell, it has units of $\n{M^{-1} \ s^{-1}}$ and not
	just $\n{s^{-1}}$. This rate equation gives the mean mRNA concentration
	$\langle\n{[mRNA]}\rangle = \frac{r_{\n{m}}}{\gamma_{\n{m}}}
	p_{\n{active}}$, which we will need in the next paragraph. The mean can also be found from the full chemical master equation, which is provided in~\nameref{S1_Text}.

@@

	\subsection*{Deterministic phase portraits}
	A standard way to analyze dynamical systems deterministically is
	to plot phase portraits. In the following, we present such plots where the
	state variables are the mRNA, the protein, and the xanthosine concentration. Note that for the low copy numbers that can occur in our system, a deterministic analysis is not necessarily meaningful. Nevertheless, in our case we find that stochastic simulations agree well with the deterministic results. Thus, deterministic phase portraits are a valid starting point.

@@

	We point out that we have not observed any oscillations in the
	system. Intuitively, they might be expected when the XapA rate is
	significantly larger than the XapB rate, but it turns out that oscillations
	cannot be obtained. Why they do not occur can be understood when looking at the regions that
	are bounded by all three nullclines: on these boundaries, the streamlines
	point into the bounded regions, so deterministically, they serve as
	trapping regions from which the system cannot escape. Once inside, the only
	possible trajectory is non-oscillatory flow towards the stable fixed point.

@@

	\paragraph*{The roles of XapA and XapB.}
	While it is clear that the bistability in the model system is due to the
	feedback loop from XapA and XapB, it is not intuitively clear if both XapA and XapB
	are necessary. The model implies that the bistability is due to XapB only. XapA
	is neither sufficient nor necessary and, within the estimated parameter
	regime, does not even have a significant influence on the system. This can
	be seen from the plots in Fig~\ref{fig6:xapAB}.
	Degradation of xanthosine by XapA lowers the xanthosine and protein concentration at the upper
	fixed point by a small amount and could, in principle, thereby make the
	high-expression solution vanish. For our choice of all other parameters,
	bistability only vanishes for $k_{\n{\alpha}} > 10^4$ which is far from what has been
	measured. However, a higher effective rate could, in principle, be achieved by
	different translation rates of XapA and XapB (see simplifications of the
	model in~\nameref{S1_Text}). Hence, we cannot exclude the possibility that
	XapA becomes so strong that it makes bistability impossible, but this is an
	extreme case. XapB, on the other hand, is essential; without it the system
	only has the one fixed point at low expression.\footnote{
		Seeger et.\ al.\cite{Seeger1995} observed that $\Delta xapB$ mutants could
		survive, but grew extremely slowly, with xanthosine as the only carbon
		source. This suggests that low-affinity
		import of xanthosine by NupC and NupG is sufficient to sustain slow growth,
		but insufficient to serve as a stand-in for XapB.
		Fig~\ref{fig6:xapAB} supports this supposition: in our model,
		with \textit{xapB} removed, the switch never activates and the cells are
		forced to survive with an extremely meager quantity of XapA to metabolize
		the abundant xanthosine.}

@@

	\paragraph*{The role of cooperativity.} 
	The model has two (putatively) cooperatively interacting binding sites for XapR on the
	\emph{xapAB} promoter and two cooperative binding sites on XapR for
	xanthosine. It is interesting to consider whether the cooperativity is
	a necessary feature for bistability. This question is motivated by the importance of
	cooperativity in ``typical'' genetic switches\cite{Gardner2000,Cherry2000}.

@@

	\paragraph*{Bimodality and the extracellular xanthosine concentration.}
	The stochastic approach results in the same bimodal distributions that were already
	seen in the deterministic investigation and experimental studies.
	Fig~\ref{fig8:stochC} shows the distribution of protein expression found in
	the simulations for different values of the extracellular xanthosine
	concentration. The parameters that were used are the same as in the previous
	section (listed in Table~\ref{table1:nondim}). However, we also find good agreement of the simulations and the deterministic results for other sets of parameters. To obtain the distributions,
	we ran the simulation 5000 times for a simulated time of $10^6 \unit{s}$
	each and started at an mRNA, protein and intracellular xanthosine count of 0.

@@

	The results agree very well with the deterministic fixed points and
	experiments: the mean numbers of mRNA, protein, and xanthosine in the
	stochastic results are as predicted from the phase portraits. It does,
	however, become clear that the phase portraits do not tell whether the cells
	will actually populate both the high and the low expression state,
	because they do not show the effective barrier height between the two states.
	In Fig~\subref{fig8:stochC}{(A)}, a deterministically bistable scenario is shown where the cells never switched to the high
	expression state during the run time of our simulations. This result
	implies that bistability, characterized by two separate
	\emph{stable} steady-states, largely persists even in the presence
	of stochastic fluctuations, suggesting that the deterministic
	picture is remarkably effective. In other words,
	the fact that bimodality only occurs with some fine tuning of
	parameters means that the circuit is a strong switch and the
	deterministic picture is sufficient except in a small region of
	parameter space. Whether or not the system is actually in that
	region of parameter space remains a question for future experiments.

@@

	\section*{Acknowledgments}
	We thank Patrick Lenggenhager for his help with \emph{Mathematica} issues, Jane Kondev and Jin Wang for interesting discussions, and Nigel Orme for assistance with illustrations.
	Portions of the experiments reported here were carried out at the
	Physiology Course at the Marine Biological Laboratory in Woods
	Hole, MA, operated by the University of Chicago.
	This work was supported by 1R35 GM118043-01 Maximizing Investigators'
	Research Award (MIRA) (to R.P.), and the Werner Siemens Foundation
	through the Swiss Study Foundation (K.S.L.). This material is based upon
	work supported by the National Science Foundation Graduate Research
	Fellowship under Grant No.\ DGE-1745301 (M.J.M.).
